---
title: "Visualizing State-Level Data"
author: "Casey Giordano"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: Cayman
    highlight: github
    self_contained: no
---
# Introduction
Let's take a look at variation in state-level data. Specifically, we are interested in the distributions of IQ scores, gross state product, health, crime, and government effectiveness. But, rather than viewing boring old tables, why don't we use some visualizations to illustrate differences? 

The data contained within this document are drawn exclusively from a published dataset. 

- McDaniel, M. A. (2006). Estimating state IQ: Measurement challenges and preliminary correlates. *Intelligence, 34(6)*, 607-619.  

```{r Load Packages, message = FALSE, echo = FALSE, warning = FALSE}

## Loading packages:
  ## If you do not have the "pacman" package, it will download it
if (!require("pacman")) install.packages("pacman")

## "pacman" is then used to load all other packages
  ## If you do not have a package, "pacman" installs it automatically
pacman::p_load(googleVis, datasets, knitr, prettydoc, tidyverse, corrplot)

## Change the options to plot the charts in an HTML doc
op <- options(gvis.plot.tag = 'chart')
```


```{r Define Functions, echo = FALSE}
## Wrapper function to easily plot the data
ChartingData <- function(DataFrame, 
                         Data, 
                         Color) {
  ## Purpose:   Plot the data using googleViz package
  ##
  ## Args: 
  ##            DataFrame: Data frame containing states and data to be plotted
  ##                         Note: Make sure states are the first column
  ##            Data:      Character value indicating which column to plot
  ##            Color:     Character or Hex to vizualize the data
  ## 
  ## Output: 
  ##            A plot
  
  ## Object to color the chart using their weird convention
  ColorValue <- paste("{colors:[ '", Color, "' ]}", sep = "")
  
  ## Create the chart itself
  Chart <- gvisGeoChart(data        = DataFrame,
                        locationvar = as.character(names(DataFrame)[1]),
                        colorvar    = as.character(Data),
                        options     = list(region      = "US",
                                           displayMode = "regions",
                                           resolution  = "provinces",
                                           wideth      = 600, 
                                           height      = 400,
                                           colorAxis   = ColorValue))
  
  ## Print out the chart
  print(Chart, "chart")
}## END function
```


```{r Load Data, warning = FALSE, echo = FALSE, message = FALSE}
## Load data
## IQ Scores by State, alphebetical order
IQ <- c( 95.7,  99.0,  97.4,  97.5,  95.5, 101.6, 103.1, 100.4,  98.4,  98.0, 
         95.6, 101.4,  99.9, 101.7, 103.2, 102.8,  99.4,  95.3, 103.4,  99.7, 
        104.3, 100.5, 103.7,  94.2, 101.0, 103.4, 102.3,  96.5, 104.2, 102.8,  
         95.7, 100.7, 100.2, 103.8, 101.8,  99.3, 101.2, 101.5,  99.5,  98.4, 
        102.8,  97.7, 100.0, 101.1, 103.8, 101.9, 101.9,  98.7, 102.9, 102.4)

## Gross State Product, alphebetical order
GSP <- c(.028, .047, .032, .027, .040, .041, .049, .059, .032, .036, .036, .029, 
         .039, .034, .034, .033, .030, .031, .030, .037, .046, .035, .040, .024, 
         .033, .026, .036, .039, .037, .044, .029, .043, .036, .032, .034, .028, 
         .033, .035, .035, .030, .034, .033, .036, .032, .032, .040, .039, .025, 
         .035, .041)

## Health Scores by state, alphebetical order
Health <- c(-11.9,  -0.8, -1.1, -14.3, 4.9,   12.5, 15.6,  -2.3, -9.9,  -9.5, 
             15.1,   8.6,  0.3,   0.5, 14.5,   7.2, -7.8, -20.5, 14.3,  -0.9, 
             16.9,   0.8, 23.3, -20.7, -3.3,   4.0, 11.2,  -5.5, 22.2,   9.0,   
              7.5,  -0.4, -5.6,  14.5,  1.8, -11.2,  7.9,   3.2, 11.6, -15.3,  
              8.9, -13.8, -4.6,  18.2, 19.8,   6.8, 11.5, -10.1, 12.7,   4.3)

## Violent crimes, alphebetical order
Crime <- c(.004, .006, .005, .005, .006, .004, .003, .006, .007, .005, .003, 
           .002, .006, .003, .003, .004, .003, .006, .001, .007, .005, .005, 
           .003, .003, .005, .003, .003, .006, .002, .004, .007, .005, .005, 
           .001, .003, .005, .003, .004, .003, .008, .002, .007, .006, .002, 
           .001, .003, .003, .003, .002, .003)

## Government Effectiveness Ratings, alphebetical order
Govt <- c(1.3, 3.3, 4.3, 3.0, 2.7, 4.0, 3.0, 6.7, 4.7, 5.0, 2.7, 4.3, 5.0, 4.3, 
          6.3, 5.7, 6.7, 5.3, 4.3, 6.3, 4.3, 7.3, 6.3, 4.0, 7.0, 4.3, 5.7, 4.0, 
          3.3, 5.0, 3.3, 3.7, 5.3, 5.0, 6.0, 3.3, 4.3, 6.3, 3.0, 6.3, 4.7, 4.7, 
          6.0, 8.0, 5.3, 7.7, 7.7, 3.7, 5.3, 3.0)

## Data Frame with the each state's raw data
  ## state.name comes from package "datasets" and already has states in 
  ## alphabetical order
RawData <- data.frame(state.name, 
                      IQ, 
                      GSP, 
                      Health, 
                      Crime, 
                      Govt)

## Convert raw data into z-scores
ZScores <- apply(X      = RawData[, -1],
                 MARGIN = 2,
                 FUN    = function(RawData) scale(x      = RawData,
                                                  center = TRUE,
                                                  scale  = TRUE))
## Turn z-scores into data frame
ZScores <- data.frame(state.name, 
                      ZScores)

## Correlation of state-level variables
CorrelationMatrix <- matrix(c(1.00, 0.28,  0.75, -0.58, 0.34,
                              0.28, 1.00,  0.34,  0.14, 0.10,
                              0.75, 0.34,  1.00, -0.66, 0.12,
                             -0.58, 0.14, -0.66,  1.00, 0.02,
                              0.34, 0.10,  0.12,  0.02, 1.00), 
                            nrow = 5, 
                            ncol = 5)

## Names of the variables in order that they appear
VariableNames <- c("IQ", 
                   "GSP", 
                   "Health", 
                   "Crime", 
                   "Govt")

## Name the column
dimnames(CorrelationMatrix) <- list(VariableNames, 
                                    VariableNames)
```

# Visualizing IQ
```{r Plot IQ, warning = FALSE, echo = FALSE, message = FALSE, results = 'asis'}
## Plots the IQ data in a green color

ChartingData(DataFrame = ZScores, 
             Data      = "IQ",
             Color     = "Green")
```

# Visualizing Gross State Product
```{r Plot GSP, warning = FALSE, echo = FALSE, message = FALSE, results = 'asis'}

## Plots GSP in a blue color
ChartingData(DataFrame = ZScores,
             Data      = "GSP",
             Color     = "Blue")
```

# Visualizing Health
```{r Plot Health, warning = FALSE, echo = FALSE, message = FALSE, results = 'asis'}

## Plots the health data in a teal color
ChartingData(DataFrame = ZScores,
             Data      = "Health",
             Color     = "teal")
```

# Visualizing Crime
```{r Plot Crime, warning = FALSE, echo = FALSE, message = FALSE, results = 'asis'}

## Plots the crime data in a red color
ChartingData(DataFrame = ZScores,
             Data      = "Crime",
             Color     = "red")
```

# Visualizing Government Effectiveness
```{r Plot Govt, warning = FALSE, echo = FALSE, message = FALSE, results = 'asis'}

## Plots the government effectiveness data in a maroon color
ChartingData(DataFrame = ZScores,
             Data      = "Govt",
             Color     = "Maroon")
```

# Putting it all together
The visualizations above show the variation of state-level variables but not the covariation. How correlated are these variables?


```{r Correlation Visualization, echo = FALSE}

## Makes a visualizations of the correlations rather than viewing raw numbers
corrplot(corr   = CorrelationMatrix,
         type   = "upper",
         diag   = FALSE,
         tl.col = "black")
```


# Raw data used in the visualizations
## Z-scores for each state
```{r Z-Score Tables, echo = FALSE}

## Table the raw data for people that want to see it
kable(x         = ZScores, 
      caption   = "State-level variables", 
      digits    = 2,
      col.names = c("State", 
                    "IQ", 
                    "GSP", 
                    "Health", 
                    "Crime", 
                    "Gov't"))
```

## Intercorrelations among state variables
```{r Correlation Matrix, echo = FALSE}

## Table the correlation matrix
kable(x       = CorrelationMatrix,
      caption = "Intercorrelations among state-level variables",
      digits  = 2)
```

